---
title: "Palmer Penguins Initial Analysis"
author: "Suzan Taha"
format: html
editor: visual
execute: 
  keep-md: true
---

## Palmer Penguin Analysis

This is an analysis of the Palmer's Penguin dataset.

## Loading Packages and Datasets

Here we will load the tidyverse package and penguins data.

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

#Read the penguins_samp1 data file from github
penguins <- read_csv("https://raw.githubusercontent.com/mcduryea/Intro-to-Bioinformatics/main/data/penguins_samp1.csv")
#See the first six rows of the data we've read in to our notebook

penguins %>% 
  head() %>%#if you want a certain number of rows -- add a number in the paranthases
  kable() %>%
  kable_styling(c("striped","hover"))
```

In the data above we are looking at a data set of penguins. This data set tells us the species of the penguins, which island they are originated from, their bill length and bill depth, flipper length, the mass and sex of each penguin as well as the year they were born.

## About our Data

The data we are working with is a data set on Penguins, which includes 8 features measured on 44 penguins. The features included are physiological features (like bill length, bill depth, flipper length, body mass, etc) as well as other features like the year the penguin was observed, the island the penguin was observed on, and the species of the penguin.

### Interesting Questions to Ask

Questions I am interested in:

-   What is the average flipper length of each species?

-   What species has the penguin with the largest bill length?

-   What species has the penguin with the largest flipper length?

-   What is the largest flipper length?

-   What is the ratio of bill length to bill depth for a penguin? What is the overall average of this metric? Does it change by species, sex, or island?

-   What is the average body mass? What about by island? By species? By sex?

-   Are there more male or female penguins? What about per island or species?

-   Does average body mass change by year?

    ### Data Manipulation

    I will be using R code to learn how to manipulate the data, specifically to filter rows, subset columns, group data, and compute summary statistics.

```{r}
penguins %>%
  count(island)
```

If we want to `filter()` and only show certain rows, we can do that too.

```{r}
#we can filter by sex (categorical variables)
penguins %>%
  filter(species == "Chinstrap")

#we can also filter by numerical variables
penguins %>%
  filter(body_mass_g >= 6000) #gives us penguins with a body mass of at least 6000grams

penguins %>%
  filter((body_mass_g >= 6000) | (island == "Torgersen"))
```

### Answering Our Questions

Most of our questions involve summarizing data, and perhaps summarizing over groups. We can summarize data using the `summarize()` function and group data using `group_by()`.

Let's find the average flipper length

```{r}
penguins %>% #average for all species
  summarize(avg_flipper_length = mean(flipper_length_mm))

penguins %>% #single species avg length
  filter(species == "Gentoo") %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))
            
penguins %>% #average separated by species (grouped average)
  group_by(species) %>%
  summarize(avg_flipper_length = mean(flipper_length_mm))

```

How many of each species do we have?

```{r}
penguins %>%
  count(species)
```

How many penguins by sex?

```{r}
penguins %>%
  count(sex)
```

How many penguins of each species are female? Male?

```{r}
penguins %>%
  group_by(species) %>%
  count(sex)
```

What is the ratio of bill length to bill depth for a penguin? What is the overall average of this metric? Does it change by species, sex, or island?

We can `mutate()` to add new columns to our data set.

```{r}
penguins_with_ratio = penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm)

#average ratio
penguins %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
  summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio),
            median_bill_ltd_ratio = median(bill_ltd_ratio))
#average ratio by group 
penguins%>%
  group_by(species) %>%
  mutate(bill_ltd_ratio = bill_length_mm / bill_depth_mm) %>%
  summarize(mean_bill_ltd_ratio = mean(bill_ltd_ratio),
            median_bill_ltd_ratio = median(bill_ltd_ratio))

```

Average body mass by year

```{r}
penguins %>%
  group_by(year) %>%
  summarize(avg_body_mass = mean(body_mass_g))
```

```{r}
penguins %>%
  count(island, species) %>%
  pivot_wider(names_from = species, values_from = n, values_fill = 0) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover","stripped"))
```

This code block is summarizing our data and finding the mean, median, sd, etc.

```{r}
penguins %>%
  summarize(
    mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
    first_quartile_bill_length =quantile(bill_length_mm, 0.25, na.rm = TRUE),
    median_bill_length = median(bill_length_mm, na.rm = TRUE),
    min_bill_length = min(bill_length_mm, na.rm = TRUE),
    third_quartile_bill_length = quantile(bill_length_mm, 0.75, na.rm = TRUE),
    standard_deviation_bill_length = sd(bill_length_mm, na.rm = TRUE)
  ) %>%
  pivot_longer(cols= everything())

```

Creating a table with columns of our choice.

```{r}
penguins %>%
  select(species, island, sex, year) %>%
  filter(species == "Chinstrap")
```

```{r}
chinstraps <- penguins %>%
  select(species, island, sex, year) %>%
  filter(species == "Chinstrap") %>%
  select(-species)

chinstraps %>%
  head()
```

Comparing mean bill depth and standard deviation per species.

```{r}
penguins %>%
  group_by(species) %>%
  summarise(
    mean_bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE),
    sd_bill_depth_mm = sd(bill_depth_mm, na.rm = TRUE),
  ) #gives us a mean and sd avearage for each species bill depth
```

### Data Visualization

-   What is the distribution of penguin flipper length?

-   What is the distribution of penguin species?

-   Does the distribution of flipper length depend on the species of penguin?

-   How many penguins were observed per year?

-   Is there any correlation between the bill length and the bill depth? \[scatter plot\]\\

    Discussion: In the graph bar plot below we are looking at how many penguins per species were observed.

    ```{r}
    penguins %>%
      ggplot() + 
      geom_bar(mapping = aes(x=species))+
      labs(title = "Counts of Penguin Species",
           x = "Species", y="Count")
    ```

    This bar plot depicts the count of each penguins species that were observed. Looking at this diagram we can see that Gentoo's take over more than 30 of the 44 penguins where as only two Chinstrap penguins were observed. This data tells us that we don't have an accurate representation of the populations of penguins.

```{r}
penguins %>%
  ggplot() +
  geom_histogram(aes(x=flipper_length_mm),
                 bins = 8,
                 fill =  "forestgreen",
                 color = "black") +
  labs(title = "Distribution of Flipper length (mm)", 
       subtitle = "Mean in black, median in blue",
       x = "Flipper Length (mm)",
       y = "" ) +
  geom_vline(aes(xintercept = mean(flipper_length_mm)), lwd = 2, lty= "dashed") +
  geom_vline(aes(xintercept = median(flipper_length_mm)), lwd = 2, lty= "dotted", color = "blue")
             #more bins = more detail, aes goes back into the data set to find info 
```

This bar plot shows the flipper lengths of our data set, outlining the mean (black) and median (blue). We can see that our mean is around 211mm whereas our median is around 215mm. The difference indicates that there may be penguins that were observed who have a relatively small flipper length.

We will now look at the distribution of species.

```{r}
penguins %>%
  ggplot()+
  geom_bar(mapping = aes(x=species), color = "black", fill = "blue") +
  labs(title = "Counts of Penguin Species",
       x = "Species", y="Count")
```

Discussion: This bar plot depicts how many penguins of each species were observed in this dataset. The majority of these penguins are Gentoo's and we only see 2 Chinstraps that were observed, creating a data set that does not define the whole population of penguins.

```{r}
penguins %>%
  ggplot() +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  labs(title = "Ratio of Bill Depth and Bill Length",
             x= "Bill Length", y ="Bill Depth") +
  geom_smooth(aes(x = bill_length_mm, y = bill_depth_mm, color = species), method = "lm")
```

Discussion: This scatter plot answers the question of whether bill length is correlated to bill depth. This diagram shows three different lines, one for each of the species. We can see that the majority of penguins observed are the Gentoo's which leads us to believe that their line of best fit is more accurate than the other species of penguins that were observed.

```{r}
penguins %>%
  ggplot() +
  geom_bar(mapping = aes(x = island, fill = species)) +
  labs(title = "Species by Island",
       x = "Island",
       y = "Species")
```

Discussion: This bar plot depicts how much of each species we observed on each of the three islands. It is noted that all Gentoo's are found on Biscoe whereas all Chinstraps are found on the island Dream. The species Adelle are found on Biscoe as well as Torgersen.

```{r}
penguins %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = bill_depth_mm, y = species)) +
  labs(title = "Bill Depth by Species",
       x = "Bill Depth (mm)",
       y = "")
```

Discussion: This bar plot explains the bill depth of each species in mm while giving averages for each of the species as well. We can see that Gentoo's have a smaller average in bill depth where Chinstraps and Adelles are closer to one another.

### A Final Question

This chunk of R code tells shows us the confidence interval for mean bill lengths.

```{r}
penguins %>%
  summarize(avg_bill_length = mean(bill_length_mm))

t.test(penguins$bill_length_mm, alternattive = "greater", mu=45, conf.level = 0.95)
```

The average bill length for a penguin that was given from our observations is about 46 mm. This average is defined by only the subset we have observed (44) and cannot be used for the whole population. The data is inaccurate in having only two Chinstrap penguins and a load ful of Gentoo's, therefore we can say this dataset does not to a good job at portraying the entire population of penguins in the world.
